### 2023-04-16 tex/article.tex
Update article to production version

% \begin{keypoints}
% \item Over SST fronts in the Gulf Stream region \al{chl} is enhanced by up to a factor two and the bloom onsets earlier by up to two weeks.
% \item The enhancement intensity increases significantly with the strength of the front.
% \item The net contribution of fronts to \al{chl} enhancement over the Gulf Stream region is less than 5\%.
% \end{keypoints}

\begin{articleBlock}{Abstract}
  Fronts affect phytoplankton growth and phenology by locally reducing stratification and increasing vertical nutrient supply.
  Biomass peaks at fronts have been observed in-situ and linked to local nutrient upwelling, and reduced stratification over fronts has been shown to induce earlier blooms in numerical models.
  However observation of these biophysical interactions through satellite imagery have been scarce, despite the opportunity to quantify them at synoptic scales.
  Here we used twenty years of SST and \al{chl} satellite data in a large region surrounding the Gulf Stream to quantify the impact of fronts on phytoplankton in contrasting regimes, from oligotrophy to bloom, and throughout the year.
  We computed an heterogeneity index HI from SST, and used it to sort fronts into weak and strong fronts based on HI thresholds.
  We observed that the localization of strong fronts corresponded to western boundary current fronts, and weak fronts to more ephemeral submesoscale fronts.
  We compared \al{chl} distributions over strong fronts, weak fronts and outside of fronts.
  We assessed three metrics, the local enhancement of \al{chl} over fronts, the global enhancement of \al{chl} due to fronts at the scale of the region, and the lag in spring bloom onset due to fronts.
  We found that weak fronts lead to a local enhancement of \al{chl} weaker than strong fronts, but because they are also more frequent they contribute equally to the regional \al{chl} budget.
  We also find the the local enhancement of \al{chl} was two to three times larger for the spring bloom than in the oligotrophic subtropical gyre.
  We also provide observational evidence that blooms start earlier over fronts, by one to two weeks.
  Nevertheless our results suggest that the spectacular impact of fronts at the local scale may be misleading, considering their impact on a regional scale budget remains limited.
\end{articleBlock}

\begin{articleBlock}{Plain Language Summary}
  Phytoplankton vary in abundance in the ocean over of large regions, and with the seasons, but also because of small-scale heterogeneities in surface temperature, called fronts, where phytoplankton growth can be favoured.
  Here, using satellite imagery, we found that fronts enhance phytoplankton much more where it is already growing well, but despite large local increases the  enhancement for the region is modest (\pct{5}). We also found that blooms start by one to two weeks earlier over fronts.
\end{articleBlock}

@@

Phytoplankton form the basis of marine food webs and are key players in the ocean carbon cycle.
The transport of limiting nutrients to the sunlit euphotic layer by advective and convective processes, and the amount of light received by the cells \encadra{which is closely related to the stratification of the water column} are two important factors that control their growth.
As there are marked contrasts in nutrient and light availability in the ocean, it follows that the global ocean can be divided into different regional biomes (or bioregions), characterized by different phytoplankton abundances and seasonality (\cite{longhurst_2007, vichi_2011a, bock_2022}).
The contrasts between biomes are largely explained by consistent physical forcings and environmental conditions, operating at the biome scale, which determine how the two main controlling factors, nutrient and light, limit growth.
For example, subtropical gyres are areas where wind-driven circulations induce a deepening of the thermocline and nutricline, resulting in oligotrophic biomes where productivity is relatively constant and low throughout the year; at higher latitudes, the nutricline is shallower and the strong seasonality of the vertical mixing will induce a multi-stage operation, with a time of reduced productivity and convective nutrient supply in winter when the mixing is strong, and a bloom in spring when the stratification sets in~(\cite{wilson_2005, williams_2011}).

In addition to these large-scale patterns, there has been considerable evidence over past years that submesoscale motions actively influence the nutrient and light environments at the horizontal scale of the order of~\qtyrange{1}{50}{\km} (\cite[see reviews by][]{levy_2012, mahadevan_2016, levy_2018}).
Submesoscale motions arise dynamically through advective interactions involving mesoscale strain that continuously create sharp submesoscale density fronts, or at more steady wind-driven density fronts, such as western boundary currents (\cite{thomas_2008, mcwilliams_2016, mahadevan_2020}).
These fronts are characterized by an energetic secondary vertical circulation, with upwelling on the warm side of the front, and downwelling on the cold side.
Submesoscale dynamics also involve restratification and suppression of vertical mixing at the front (\cite{thomas_2008a}).
Thus submesoscale dynamics occurring at ocean fronts may affect phytoplankton in various ways.
Of interest here, the upward branch of the secondary circulation may enhance phytoplankton growth by transporting nutrients into the euphotic zone, while the downward branch may subduct biomass and excess nutrients into the subsurface (\cite{calil_2011, omand_2015, hauschildt_2021}).
In addition, in highly seasonal regimes where productivity is slowed in winter due to deep mixing, submesoscale restratification may promote localized phytoplankton blooms before the large-scale outburst associated with seasonal stratification~(\cite{mahadevan_2012}).

Despite numerous local observations and a strong theoretical basis for these processes (\cite[e.g.\ recent studies by][]{marrec_2018, little_2018, verneil_2019, ruiz_2019, uchida_2020, kessouri_2020, tzortzis_2021}), their integrated contribution at the scale of regional biomes is still largely unknown.
Indeed, as most fronts form, move and dissipate continuously on time scales of days to weeks, they are particularly difficult to sample, and this limitation is reinforced by the fact that only a limited number of fronts can be observed with in situ field observations.
Thus satellite-derived estimates of \al{chl} (hereafter \as{chl}) although limited to the surface and an imperfect proxy for phytoplankton biomass, are the only data that allow to track the impact of fronts synoptically over large areas.
A first attempt to assess the contribution of fine scales to regional satellite \as{chl} budgets was based on a geostatistical analysis derived from data at 9km resolution (\cite{doney_2003}), extended later in \textcite{glover_2018}, with which they examined the change in spatial variance with distance.
This methodology was too coarse to reveal the impact of submesoscales but confirmed the important role of mesoscale eddies in stirring large scale gradients of phytoplankton abundance.
The role of submesoscales has been assessed with three different methods.
\textcite{guo_2019} combined ocean color data with altimetry and drifting floats, and estimated that, over subtropical gyres of the global ocean, the respective contributions of mesoscales and submesoscales to high \as{chl} anomalies were comparable in magnitude.
\textcite{keerthi_2022} proposed an approach based on deconvolution of local \as{chl} time series into different timescales; they observed that sub-seasonal time scales contributed roughly~\pct{30} of the total satellite \as{chl} variance and were associated with small~(\qty{<100}{\km}) spacial scales.
Finally, the most quantitative approach was proposed by \textcite{liu_2016}, which they applied to the North Pacific Subtropical Gyre.
They detected sea-surface temperature (SST) fronts by computing an index that measures the local heterogeneity of the SST field from satellite SST data.
This allowed them to compare satellite \as{chl} values over areas impacted by fronts (characterized by a large value of the heterogeneity index) with values over areas that were not impacted.
They found that the increase in \as{chl} over the fronts was negligible in summer but reached almost~\pct{40} in winter.

Here we built on this last approach, and quantify the excess \as{chl} due to the presence of fronts at the scale of biomes.
This more global quantification will depend on several factors, since the local contribution of fronts to \as{chl} depends on many factors.
First it depends on how efficient fronts are at supplying nutrients, which itself depends on how deep the fronts reach into the nutricline, and of the seasonality of this supply.
It should be noted that the overall efficiency of fronts has been questioned (\cite{levy_2018}); firstly because submesoscale fronts are more numerous in winter when convective nutrients input is also greatest; and secondly because the submesoscale vertical currents are often trapped within the mixed-layer, and may not reach the nutricline, which is often well below the base of the boundary mixed-layer.
Second, the overall contribution of fronts will differ between biomes, with submesoscale vertical advection of nutrients likely to be more important in oligotrophic biomes where other nutrient supply routes are scarce, and submesoscale restratification in blooming biomes.
Finally, the contribution of fronts will depend on their spatio-temporal footprint, which also varies seasonally (\cite{callies_2015}) and regionally~(\cite{mauzole_2022}).

Thus, more precisely, our intention is to explore and quantify how the contribution of fronts to biome-scale \as{chl} varies in three contrasted biomes, ranging from subtropical to subpolar, varies along the year, and varies with the occurrence and strength of fronts.
We focus our analysis on the North-Atlantic region surrounding the Gulf Stream, where multiple biomes and fronts of different strengths are found in a limited geographical area (\cite{bock_2022}) with strong seasonality.
In the south, our study area encompasses part of the North Atlantic subtropical gyre, characterized by an oligotrophic regime, year-long low productivity.
In the north, north of the Gulf Stream jet, is a more productive subpolar regime characterized by a recurrent spring bloom.
In between, there is a moderately productive regime, with maximum productivity in winter.
Another feature that makes this study area particularly relevant is that it has different types of fronts.
On the one hand, there are two strong persistent fronts, the Gulf Stream and the shelf-break front, which are held in place by topography and atmospheric circulation, and which are both associated with strong and deep-reaching vertical circulations (\cite{flagg_2006, liao_2022}).
But there are also more ephemeral and weaker fronts, created by mesoscale strain that are continuously forming at more random locations (\cite{drushka_2019, sanchez-rios_2020}).
We use satellite data of \as{chl} and SST, and extend on the approach of \textcite{liu_2016}, to distinguish between persistent and ephemeral fronts.
We evaluate the impact of both types of fronts on \as{chl} on the basis of three indicators, the excess (or deficit) \as{chl} over fronts at the local scale of the front, the surplus \as{chl} attributable to fronts at the scale of regional biomes, and the change in the timing of the \as{chl} spring bloom over fronts.

@@

Our approach combines daily satellite SST data, which are used to detect fronts and sort them by their strength, with daily satellite surface \as{chl}, from which we derive anomalies over fronts.
Our region of interest is the North Atlantic from~\latlon{15N} to~\latlon{55N}, and from~\latlon{40W} to the North American shelf break (\nref{fig:zone-separation}).
This region covers three biomes, a more oligotrophic one in the south, a more productive one in the north where a spring bloom occurs, and an intermediate biome between the two, described below.
We will refer to them in the following as permanent subtropical biome (PSB), seasonal subtropical biome (SSB) and subpolar biome (PB), moving from south to north.
They are separated from south to north by a fixed boundary at~\latlon{32N} and by a variable boundary near~\latlon{40N} dynamically set at each time step to the instantaneous position of the Gulf Stream.

@@

For \as{chl}, we used the \abbrv{L3} product distributed by ACRI-ST over the period 2000--2020, generated by Copernicus-GlobColour, constructed with data from different sensors (\as{seawifs}, \as{modis} Aqua \& Terra, \as{meris}, \as{viirs}-\as{snpp} \& \as{jpss}[1], \as{olci}[\abbrv{-S3A}] \& \abbrv{S3B}) merged and reprocessed, available daily at 4km resolution\footnote{Dataset \dataname{chl_globcolour} décrit~\nref[section]{sec:donnees-chl_globcolour}}.

For SST, we used the European Space Agency Sea Surface Temperature Climate Change Initiative analysis product version 2.1 (\cite{merchant_2019, good_2020})\footnote{Dataset \dataname{sst_esacci} décrit \nref[section]{sec:donnees-sst_esacci}}, also available daily at~\qty{4}{\km} resolution over the period 2000--2020.
This product combines data from all available infrared sensors (\as{atsr}, \as{slstr}, and \as{avhrr} sensors), ensuring good resolution where data are available, unlike other SST products which also include microwave and in-situ measurements, resulting in considerable smoothing of the SST field. %chktex 36
Where SST data is not available, spatial interpolation is performed to obtain a cloud free product which, at the cost of resolution on finer features, provides complete synoptic coverage of our large study area.
This interpolation tends to provide an underestimate of the detection of fronts, as the SST field is smoother over cloud-covered areas (\cite{merchant_2019}).
However, the combination of several sensors allows to reduce these areas to a minimum.
Furthermore, we have only considered cloud-free pixels for our analysis, which ensures that cloudy areas are not taken into account in our quantification.

@@

The three biomes in our domain correspond to well known production regimes which have been described previously, although sometimes with different names.
The region is characterized by the presence of a large-scale north-south gradient in \as{chl}.
All pixels where water depth is less than~\qty{<1500}{\m} (red isobath in \nref[figure]{fig:zone-separation}) are masked to exclude the continental shelf.

The permanently oligotrophic regime to the south of our study area, characterized by warm waters and low \as{chl} (\nref{fig:zone-separation}), is known as the subtropical gyre permanently stratified biome (\cite{sarmiento_2004}) or permanent deep \as{chl} maximum biome (\cite{bock_2022}).
Their is no clear physical boundary to the northern limit of this permanent subtropical biome, so we have chosen the latitudinal limit of~\latlon{32N} to delineate it, which roughly corresponds to the~\qty{0.1}{\mgm}~\as{chl} isocontour in annual mean \as{chl}.

North of~\latlon{32N}, the seasonal subtropical biome is also mainly oligotrophic, with intermediate levels of \as{chl} and temperature, and characterized by slightly increased productivity in winter.
This second biome is known as the subtropical gyre seasonally stratified biome (\cite{sarmiento_2004}).
It is bounded to the north by the meanders of the Gulf Stream jet.
The Gulf Stream jet conveys warm, salty waters poleward along the Florida coast up to Cape Hatteras~(\latlon{35N}), where the jet separates from the continental shelf and meanders essentially zonally.
The north wall of the jet, so called because of its steep temperature gradient, marks the sharp, sinuous and unsteady northern limit of this biome.

To the north of the Gulf Stream is the Slope sea which extends to the shelf-break, with colder and fresher waters (\cite{linder_1998}).
Aligned with the shelf break, a persistent front with an intensified surface jet separates the shelf waters (excluded from this study) from the slope sea.
This highly productive subpolar biome, known as subpolar waters (\cite{sarmiento_2004}) and high-chlorophyll-bloom (\cite{bock_2022}), is characterized by a strong spring bloom whose onset is tied to the spring stratification of the mixed-layer.

The position of the north wall of the Gulf stream, that delimits the subpolar and seasonal subtropical biomes (black meandering contour \nref{fig:zone-separation}a-b), is determined at each daily time step by thresholding the daily SST map.
The daily threshold values (black vertical line \nref{fig:zone-separation}c) are determined from the daily SST distributions above \latlon{32N}~(blue line, continued by the yellow line \nref{fig:zone-separation}c).
Indeed, the Gulf Stream is easily identifiable in this distribution as it is manifested by a temperature peak (yellow line \nref{fig:zone-separation}c).
To detect the start of this peak, which marks our boundary, we fit the peak with a Gaussian, and define the threshold temperature as the mean minus twice the standard deviation.
Then, the threshold temperature time series is median filtered over an 8-day window to eliminate spurious detection anomalies.
This separation method ensures quasi-unimodal \as{chl} distributions within each biome (\nref{fig:zone-separation}d).

With the above delimitation of biomes, the two atmospherically and topographically fronts, \ab{ie} the Gulf stream and the shelf-break, are located within the seasonal subtropical biome and subpolar biome respectively, while the permanent subtropical biome only contains submesoscale fronts.

\begin{figure}
  \centering
  \includegraphics[width=12cm]{article/zone_separation.pdf}
  \caption[Delimitation of biomes]{
    Delimitation of the three biomes in the Gulf stream extension region: the Permanent Subtropical Biome~(PSB, south of the dashed line at~\latlon{32N}), the Seasonal Subtropical Biome~(SSB, between \latlon{32N} and the meandering Gulf stream northern wall on that day marked with the black contour), and the Subpolar Biome~(PB, north of the Gulf stream northern wall).
    (a)~SST and (b)~\as{chl} snapshots on the 22 April 2007 (with data masked by clouds in white), (c)~SST and (d)~\as{chl}  distribution within each biome for the same day (PB:\@ blue, SSB:\@ yellow, PSB:\@ red).
    The black line in~(c) shows the SST threshold value detected to delimit the Gulf Stream northern wall~(see methods section).
    The x"~axis of the distributions correspond to the x"~axis scale of the corresponding color bars.
    The red line follows the~\qty{1500}{m} isobath.
    Data on the continental shelf~(\qty{<1500}{\m}) is not considered here and have been masked.
  }%
  \label{fig:zone-separation}
\end{figure}

@@

Our front detection method aims to sort the domain into pixels that are located (or not) over fronts, at each daily time step.
We follow the approach of \textcite{liu_2016} that builds upon the well established Cayula-Cornillon algorithm (\cite{cayula_1992, belkin_2009}), and identify pixels as belonging to fronts when the region in the vicinity of the pixel is characterized by high SST heterogeneity.
More precisely, on each pixel and for each time step, we compute an Heterogeneity Index~(HI) which quantifies the heterogeneity of the SST in a square window comprising a few pixels~(\numproduct{7 x 7}) centered on the pixel of interest.
Thus HI measures the SST heterogeneity at spatial scales less than~\qty{30}{\km}.
The HI is defined as the weighted sum of the skewness~\ab{skew}, standard deviation~\ab{std}, and bimodality~\ab{bimod} of SST within the window (\(\am{hi} = a \left( b \am{skew} + c \am{std} + d \am{bimod} \right)\), with \(a\), \(b\), \(c\), \(d\),~constant normalization coefficients).

>>>>

% \begin{keypoints}
% \item Over SST fronts in the Gulf Stream region \al{chl} is enhanced by up to a factor two and the bloom onsets earlier by up to two weeks.
% \item The enhancement intensity increases significantly with the strength of the front.
% \item The net contribution of fronts to \al{chl} enhancement over the Gulf Stream region is less than \pct{5}.
% \end{keypoints}

\begin{articleBlock}{Abstract}
  Fronts affect phytoplankton growth and phenology by locally reducing stratification and increasing nutrient supplies.
  Biomass peaks at fronts have been observed in-situ and linked to local nutrient upwelling and/or lateral transport, while reduced stratification over fronts has been shown to induce earlier blooms in numerical models.
  Satellite imagery offers the opportunity to quantify these induced changes in phytoplankton over a large number of fronts and at synoptic scales.
  Here we used twenty years of Sea Surface Temperature (SST) and \al{chl} (\as{chl}) satellite data in a large region surrounding the Gulf Stream to quantify the impact of fronts on surface \as{chl} (used as a proxy for phytoplankton) in three contrasting bioregions, from oligotrophic to blooming ones, and throughout the year.
  We computed an Heterogeneity Index (HI) from SST to detect fronts, and used it to sort fronts into weak and strong ones based on HI thresholds.
  We observed that the location of strong fronts corresponded to the persistent western boundary current fronts, and weak fronts to more ephemeral submesoscale fronts.
  We compared \as{chl} distributions over strong fronts, over weak fronts and outside of fronts in the three bioregions.
  We assessed three metrics, the \as{chl} excess over fronts at the local scale of fronts, the surplus in \as{chl} induced at the bioregional scale, and the lag in spring bloom onset over fronts.
  We found that weak fronts are associated with a local \as{chl} excess weaker than strong fronts, but because they are also more frequent they contribute equally to the regional \as{chl} surplus.
  We also found that the local excess of \as{chl} was two to three times larger in the bioregion with a spring bloom than in the oligotrophic bioregion, which can be partly explained by the transport of nutrients by the Gulf Stream.
  We found strong seasonal variations in the amplitude of the \as{chl} excess over fronts, and we show periods of \as{chl} deficit over fronts north of \latlon{45N} that we attribute to subduction.
  Finally we provide observational evidence that blooms start earlier over fronts, by one to two weeks.
  Our results suggest that the spectacular impact of fronts at the local scale of fronts (up to + \pct{60}) is more limited when considered at the regional scale of bioregions (less than +5 \%), but may nevertheless have implications for the region's overall ecosystem.
\end{articleBlock}

\begin{articleBlock}{Plain Language Summary}
  Phytoplankton vary in abundance in the ocean over large regions, and with the seasons, but also because of small-scale heterogeneities in surface temperature, called fronts.
  Here, using satellite imagery, we found that fronts enhance phytoplankton much more where it is already growing well, but despite large local increases the enhancement for the region is modest (\pct{5}).
  We also found that blooms start by one to two weeks earlier over fronts. These effects may have implications for ecosystems.
\end{articleBlock}

@@

Phytoplankton form the basis of marine food webs and are key players in the ocean carbon cycle.
The transport of limiting nutrients to the sunlit euphotic layer by advective and convective processes, and the amount of light received by the cells \encadra{which is closely related to the stratification of the water column} are two important factors that control their growth.
As there are marked contrasts in nutrient and light availability in the ocean, it follows that the global ocean can be divided into different regional biomes (or bioregions), characterized by different phytoplankton abundances and seasonality (\cite{longhurst_2007, vichi_2011a, bock_2022}).
The contrasts between biomes are largely explained by consistent physical forcings and environmental conditions (such as nutrient sources), operating at the biome scale, which determine how the two main controlling factors, nutrient and light, limit growth.
For example, subtropical gyres are areas where negative wind-stress curl induces a deepening of the thermocline and nutricline, resulting in oligotrophic biomes where productivity is relatively constant and low throughout the year.
At higher latitudes, where the wind-stress curl is positive, the nutricline is shallower; the strong seasonality of the vertical mixing will induce a multi-stage operation, with a time of reduced productivity and convective nutrient supply in winter when the mixing is strong, and a bloom in spring when the stratification sets in (\cite{wilson_2005, williams_2011}).
In addition to these large-scale patterns, there has been considerable evidence over past years that the nutrient and light environments are modified at ocean fronts, with consequences on phytoplankton (see reviews by \cite{levy_2012, mahadevan_2016, levy_2018}).

Ocean fronts are narrow zones where horizontal gradients in water properties (temperature, salinity, nutrients, etc.) are significant \citep{belkin_2009a}, and are sometimes described as discontinuities because of their abrupt nature \citep{mauzole_2022}.
\Textcite{levy_2018} distinguished two types of fronts. Persistent fronts like those associated with the Gulf Stream and Kuroshio are locked in place by the coastal boundary and large-scale atmospheric forcing.
Their forcing is directly balanced by submesoscale symmetric instability, which takes energy mostly from the kinetic energy of the jet, and baroclinic instability, which converts the potential energy of the sloping density surfaces into large meanders and eddies.
Their ephemeral cousins are continuously forming, moving, and dissipating at the ocean surface.
They are being strained by mesoscale eddies, which intensifies their geostrophic along-front currents, and feeds fluid instabilities generating submesoscale vortices and filaments.
Importantly, a front will generate a cross-frontal ageostrophic secondary circulation \encadra*{an overturning circulation directed in the direction of trying to flatten the density surfaces in the front, with upwelling on the warm side of the front, and downwelling on the cold side (\cite{thomas_2008, mcwilliams_2016, mahadevan_2020})}.
The highly energetic persistent fronts are characterized by a deep vertical velocity structure that reaches the thermocline, while ephemeral fronts have associated cross-frontal secondary circulations that are generally confined to the vertically well-mixed upper layer of the ocean.
Frontal dynamics also involves increased stratification, reduction of mixed-layer depth and suppression of vertical mixing at the front (\cite{thomas_2008a}).

Fronts may affect phytoplankton in various ways. The upward branch of the secondary circulation may enhance phytoplankton growth by transporting nutrients into the euphotic zone (\cite{johnson_2010, wilson_2021}).
The downward branch may subduct biomass and excess nutrients into the subsurface (\cite{calil_2011, omand_2015, hauschildt_2021}).
Persistent fronts may act as conduct for nutrients over large distances, known as nutrient streams (\cite{pelegri_1996, williams_2011a, long_2022}).
Advective transport along fronts may transport nutrients to the shallow flanks of subtropical gyres (\cite{letscher_2016, gupta_2022}).
Finally, in highly seasonal regimes where productivity is slowed in winter due to deep mixing, increased restratification over fronts may promote localized phytoplankton blooms before the large-scale outburst associated with seasonal stratification (\cite{mahadevan_2012}).
These effects all together are responsible for local anomalies in the distribution of phytoplankton over fronts.
Thus fronts are not only a physical boundary but also constitute specific habitats for phytoplankton (\cite{mangolte_2023}).

Despite numerous local observations and a strong theoretical basis for the physical processes affecting phytoplankton growth over fronts (e.g.\ recent studies by \cite{marrec_2018, little_2018, verneil_2019, ruiz_2019, uchida_2020, kessouri_2020, tzortzis_2021}), their integrated contribution at the scale of regional biomes is still largely unknown.
Ephemeral fronts move and dissipate continuously on time scales of days to weeks, and are thus particularly difficult to sample.
This limitation is reinforced by the fact that only a limited number of fronts can be observed with in situ field observations.
Thus satellite-derived estimates of \al{chl} (hereafter \as{chl}) although limited to the surface of the ocean and an imperfect proxy for phytoplankton biomass, are the only data that allow to track the impact of fronts synoptically over large areas.
A first attempt to assess the contribution of small scale physical processes to regional satellite \as{chl} budgets was based on a geostatistical analysis derived from data at 9km resolution (\cite{doney_2003}), extended later in \textcite{glover_2018}, with which they examined the change in spatial variance with distance.
This methodology was too coarse to reveal the impact of frontal processes, but confirmed the important role of mesoscale eddies in stirring large scale gradients of phytoplankton abundance.
The role of fronts has been assessed with three different methods.
\Textcite{guo_2019} combined ocean color data with altimetry and drifting floats, and estimated that, over subtropical gyres of the global ocean, the respective contributions of mesoscale dynamics and submesoscale frontal dynamics to high \as{chl} anomalies were comparable in magnitude.
\Textcite{keerthi_2022} proposed an approach based on deconvolution of local \as{chl} time series into different timescales; they observed that sub-seasonal time scales contributed roughly \pct{30} of the total satellite \as{chl} variance and were associated with small (\qty{<100}{\km}) spatial scales \encadra*{which included both the mesoscale and the submesoscale}.
Finally, the most quantitative approach, and the only one directly related with co-localization with fronts, was proposed by \textcite{liu_2016}, which they applied to the North Pacific Subtropical Gyre.
They detected sea-surface temperature (SST) fronts by computing an index that measures the local heterogeneity of the SST field from satellite data.
This allowed them to compare satellite \as{chl} values over areas impacted by fronts (characterized by a large value of the heterogeneity index) with values over areas that were not impacted.
They found that the increase in \as{chl} over the fronts was negligible in summer but reached almost \pct{40} in winter.

Here we build on this last approach, and quantify the surplus \as{chl} induced by fronts at the scale of biomes.
The excess \as{chl} over fronts depends on how efficient fronts are at supplying nutrients, which itself depends on how deep the fronts reach into the nutricline, on the seasonality of this vertical supply, and also on the presence of nutrient sources that are advected horizontally along the front.
The impact of fronts is also expected to differ between biomes, with submesoscale advection of nutrients likely to be more important in oligotrophic biomes where other nutrient supply routes are scarce, and submesoscale restratification in blooming biomes.
Finally, the regional surplus \as{chl} induced by fronts at the scale of biomes will depend on the spatio-temporal footprint of fronts, which also varies seasonally (\cite{callies_2015}) and regionally (\cite{mauzole_2022}).
Thus, our intention is to explore and quantify how the contribution of fronts to biome-scale \as{chl} varies in three contrasted biomes, ranging from subtropical to subpolar, varies along the year, and varies with the occurrence and strength of fronts.

We focus our analysis on the North-Atlantic region surrounding the Gulf Stream, where multiple biomes and fronts of different strengths are found in a limited geographical area (\cite{bock_2022}) with strong seasonality.
In the south, our study area encompasses part of the North Atlantic subtropical gyre, characterized by an oligotrophic regime, year-long low productivity.
In the north, north of the Gulf Stream jet, is a more productive subpolar regime characterized by a recurrent spring bloom.
In between, there is a moderately productive regime, with maximum productivity in winter.
Another feature that makes this study area particularly relevant is that it has two strong persistent fronts, the Gulf Stream and the shelf-break front, which are both associated with strong and deep-reaching vertical circulations (\cite{flagg_2006, liao_2022}), with the Gulf Stream being a recognized horizontal nutrient pathway toward the North Atlantic (\cite{pelegri_1996, williams_2011}).
But there are also plethora of ephemeral fronts continuously forming at more random locations (\cite{drushka_2019, sanchez-rios_2020}).

@@

We use satellite data of \as{chl} and SST and extend on the approach of \textcite{liu_2016} to distinguish between persistent and ephemeral fronts.
We evaluate the impact of both types of fronts on \as{chl} on the basis of three indicators, the excess (or deficit) \as{chl} over fronts at the local scale of the front, the surplus \as{chl} attributable to fronts at the scale of regional biomes, and the change in the timing of the \as{chl} spring bloom over fronts.

@@

Our approach combines daily satellite SST data, which are used to detect fronts and sort them by their strength, with daily satellite surface \as{chl}, from which we derive anomalies over fronts.
For \as{chl}, we used the \abbrv{L3}~product distributed by ACRI"~ST over the period 2000--2020, generated by Copernicus-GlobColour, constructed with data from different sensors (\as{seawifs}, \as{modis} Aqua \& Terra, \as{meris}, \as{viirs}-\as{snpp} \& \as{jpss}[\textlf{-1}], \as{olci}[\abbrv{-S3A}] \& \abbrv{S3B}) merged and reprocessed, available daily at \qty{4}{\km} resolution (\cite{garnesson_2019})\footnote{Dataset \dataname{chl_globcolour} décrit~\nref[section]{sec:donnees-chl_globcolour}}.

For SST, we used the \al{esa} \al{sst} \al{cci} analysis product version 2.1 (\cite{merchant_2019, good_2020})\footnote{Dataset \dataname{sst_esacci} décrit \nref[section]{sec:donnees-sst_esacci}}, also available daily at \qty{4}{\km} resolution over the period 2000--2020.
This product combines data from all available infrared sensors (\as{atsr}, \as{slstr}, and \as{avhrr}), ensuring good resolution where data are available, unlike other SST products which also include microwave and in-situ measurements, resulting in considerable smoothing of the SST field. %chktex 36
Where SST data is not available, spatial interpolation is performed to obtain a cloud free product which, at the cost of resolution on finer features, provides complete synoptic coverage of our large study area.
This interpolation tends to underestimate the detection of fronts, as the SST field is smoother over cloud-covered areas (\cite{merchant_2019}).
However, the combination of several sensors allows to reduce these areas to a minimum.
Furthermore, we have only considered cloud-free pixels for our analysis, which ensures that cloudy areas are not taken into account in our quantification.

@@

Our region of interest is the North Atlantic from \latlon{15N} to \latlon{55N}, and from \latlon{40W} to the North American shelf break (\nref{fig:zone-separation}).
We focus on the open ocean and exclude the continental shelf, thus all pixels where water depth is less than \qty{1500}{\km} (red isobath in \nref[figure]{fig:zone-separation}) are masked.
The region is characterized by the presence of a large-scale north-south gradient in \as{chl} and is sorted into three open-ocean biomes.

The oligotrophic Permanent Subtropical Biome (PSB) \encadra{also known as the subtropical gyre permanently stratified biome (\cite{sarmiento_2004})} to the south of our study area is characterized by warm waters and low \as{chl} (\nref{fig:zone-separation}).
There is no clear physical boundary to its northern limit so we have chosen the latitudinal limit of \latlon{32N} to delineate it, which roughly corresponds to the \qty{0.1}{\mgm} \as{chl} iso-contour in annual mean \as{chl}.
There is no persistent front in the PSB\@.

North of \latlon{32N}, the Seasonal Subtropical Biome (SSB) \encadra{also known as the subtropical gyre seasonally stratified biome (\cite{sarmiento_2004}) and as the permanent deep \as{chl} maximum biome (\cite{bock_2022})} is also mainly oligotrophic, with intermediate levels of \as{chl} and temperature, and characterized by slightly increased productivity in winter.
It is bounded to the north by the meanders of the Gulf Stream jet.
When the Gulf Stream enters the SSB, it conveys warm, nutrient rich, salty waters poleward along the Florida coast up to Cape Hatteras (\latlon{35N}), where it separates from the continental shelf and meanders essentially zonally.
The north wall of the Gulf Stream, so called because of its steep temperature gradient, marks the sharp, sinuous and unsteady northern limit of the SSB\@.

To the north of the Gulf Stream is the Slope sea which extends to the shelf-break, with colder and fresher waters (\cite{linder_1998}).
Aligned with the shelf break, a persistent front with an intensified surface jet separates the shelf waters (excluded from this study) from the slope sea.
This highly productive sub-Polar Biome (PB) \encadra{also known as subpolar waters (\cite{sarmiento_2004}) and as high-chlorophyll-bloom (\cite{bock_2022})} is characterized by a strong spring bloom whose onset is tied to the spring stratification of the mixed-layer.
The shelf-break front is thus comprised within the PB\@.

The position of the north wall of the Gulf Stream, that delimits the SSB and the PB (black meandering contour in \nref[figure]{fig:zone-separation}a-b), is determined at each daily time step by thresholding the daily SST map.
The daily threshold values (black vertical line in \nref[figure]{fig:zone-separation}c) are determined from the daily SST distributions above \latlon{32N} (blue line, continued by the yellow line in \nref{fig:zone-separation}c).
Indeed, the Gulf Stream is easily identifiable in this distribution as it is manifested by a temperature peak (yellow line in \nref[figure]{fig:zone-separation}c).
To detect the start of this peak, which marks our boundary, we fit the peak with a Gaussian, and define the threshold temperature as the mean minus twice the standard deviation.
Then, the threshold temperature time series is median filtered over an 8-day window to eliminate spurious detection anomalies.
This separation method ensures quasi-unimodal \as{chl} distributions within each biome (\nref{fig:zone-separation}d).

\begin{figure*}
  \includegraphics[width=12cm]{fig01.pdf}
  \caption[Delimitation of biomes]{
    Delimitation of the three biomes in the open ocean Gulf Stream extension region: the Permanent Subtropical Biome (PSB, south of the dashed line at \latlon{32N}), the Seasonal Subtropical Biome (SSB, between \latlon{32N} and the meandering Gulf Stream northern wall on that day marked with the black contour), and the sub-Polar Biome (PB, north of the Gulf Stream northern wall).
    (a)~SST and (b)~\as{chl} snapshots on the 22 April 2007 (with data masked by clouds in white). The red line follows the \qty{1500}{\m} isobath.
    Data on the continental shelf (\qty{<1500}{\m}) is not considered here and have been masked.
    (c)~SST and (d)~\as{chl} distribution within each biome for the same day (PB:~blue, SSB:~yellow, PSB:~red).
    The black line in (c) shows the SST threshold value detected to delimit the Gulf Stream northern wall (see methods section).
  }%
  \label{fig:zone-separation}
\end{figure*}

@@

Our front detection method aims to sort the domain into pixels that are located (or not) over fronts, at each daily time step.
We follow the approach of \textcite{liu_2016} that builds upon the well established Cayula-Cornillon algorithm (\cite{cayula_1992, belkin_2009}), and identify pixels as belonging to fronts when the region in the vicinity of the pixel is characterized by high SST heterogeneity.
More precisely, on each pixel and for each time step, we compute an Heterogeneity Index (HI) which quantifies the heterogeneity of the SST in a square window comprising a few pixels (\numproduct{7 x 7}) centered on the pixel of interest.
Thus HI measures the SST heterogeneity at spatial scales less than \qty{30}{\km}.
This spatial scale captures the heterogeneity associated with sub-mesoscale fronts and persistent fronts (which are wider than sub-mesoscale fronts), with the edges of mesoscale eddies (which are also considered as fronts), but not with the centre of eddies as their diameter is generally \qty{>100}{\km} (see for instance \textcite{contreras_2023} which nicely shows the scales of these different features in this region).
The HI is defined as the weighted sum of the skewness~\as{skew}, standard deviation~\as{std}, and bimodality~\as{bimod} of SST within the window (\(\am{hi} = a \paren{ b \am{skew} + c \am{std} + d \am{bimod} }\), with \(a\), \(b\), \(c\), \(d\), constant normalization coefficients).
